Symmetric Circuits for Rank Gates

### 2017-11-13 sections/background.tex
13-Nov-2017

\begin{definition}
  Let $S = \{s_1, \ldots , s_p\}$ be a set of sort symbols and $A = A_{s_1}
  \sqcup \ldots \sqcup A_{s_p}$ and $B = B_{s_1}\dotcup \ldots \dotcup
  B_{s_p}$ be disjoint unions of non-empty sets. A \emph{sorted bijection}
  between $A$ and $B$ is a bijection $f: A \rightarrow B$ such that $f (A_s) =
  B_s$ for all $s \in S$.
\end{definition}

@@

>>>>

\begin{definition}
  Let $S = \{s_1, \ldots , s_p\}$ be a set of sort symbols and $A = A_{s_1}
  \uplus \ldots \uplus A_{s_p}$ and $B = B_{s_1}\uplus \ldots \uplus
  B_{s_p}$ be disjoint unions of non-empty sets. A \emph{sorted bijection}
  between $A$ and $B$ is a bijection $f: A \rightarrow B$ such that $f (A_s) =
  B_s$ for all $s \in S$.
\end{definition}

@@

\subsection{Basic Notation}

Let $f : X \rightarrow Y$ be a function and $\sim$ be an equivalence relation on Y. We define $f/{\sim} : X \rightarrow Y/{\sim}$ by $f(x) = [f(x)]_{\sim}$. We call this the quotient of the function by $\sim$. For any two sets $X, Y$ we let $X^{\underline{Y}}$ denote the set of all injective functions from $Y$ to $X$.

### 2017-11-13 sections/introduction.tex
13-Nov-2017

\documentclass[../paper.tex]{subfiles}
\begin{document}
A finite relational structure, such as a graph, can be represented as a binary
string in a standard way (for instance, by enumerating its adjacency matrix).
When we speak of the computational complexity of properties of graphs, we assume
such an encoding in order to present the structure as input to a machine or
circuit or other appropriate computational model. The properties of structures
we are interested in studying are invariant under isomorphisms. This requirement
translates into an invariance condition on sets of strings---say that a language
$L \subseteq \{0,1\}^\star$ is a \emph{graph property} if, whenever $x \in L$
for a string $x$ representing a graph $G$, we have $y\in L$ for any string $y$
obtained from $x$ by re-ordering the vertices of $G$. The central open question
in descriptive complexity---whether there is a logical characterisation of the
polynomial-time decidable properties of finite relational structures--- can then
be understood as the question of characterizing those languages $L$ that are in
$\PT$ and are graph properties.

Anderson and Dawar~\cite{AndersonD17} study this question through the lens of
circuit complexity, by considering \emph{symmetric circuits}. These are circuits
that necessarily accept graph properties, with this fact explicitly witnessed by
automorphisms of the circuit. To be precise, we can think of a property of
(directed) graphs on $n$ vertices as being given by a Boolean function $f:
\{0,1\}^{n^2} \rightarrow \{0,1\}$ which takes as input the $n^2$ potential
edges and outputs $1$ if the property is satisfied by the graph given on the
input. The function $f$ is \emph{graph-symmetric} in the sense that any
permutation of its inputs induced by a permutation of the $n$ vertices leaves
$f$ unchanged (note this is not the same thing as requiring that $f$ is
invariant under all permutations of its inputs). A symmetric circuit is a
circuit computing a Boolean function of $n^2$ inputs where each permutation of
$[n]$ can be extended to an automorphism of the circuit. Such a circuit
necessarily computes an invariant function.

The results of~\cite{AndersonD17} show a tight relationship between the
decdability of a graph property by means of a polynomial-size family of
symmetric circuits and its definability in extensions of fixed-point logic. In
particular, they show that a property is definable in fixed-point logic with
counting ($\FPC$) if, and only if, it is decidable by a polynomially-uniform
family of symmetric circuits using Boolean and majority gates (or, equivalently,
threshold gates). Without the symmetry restriction, polynomial-size families of
circuits have the same computational power whether we restrict ourselves to
Boolean gates only or allow richer bases such as threshold gates. One
consequence of the results in~\cite{AD14} is that, with the requirement of
symmetry, these different bases do yield different expressive power, suggesting
some lack of robustness to the model. One of the results in the present paper
shows that the model with majority circuits is robust in the sense that it is
able to simulate, by means of symmetric circuits, any symmetric circuit composed
of gates computing Boolean functions that are themselves symmetric (in the sense
that they are invariant under \emph{all} permutations of their inputs. Thus, any
property decidable by a polynomially-uniform family of such circuits is
definable in $\FPC$.

This motivates the study of symmetric circuit families that involve gates which
themselves may compute non-symmetric functions. In particular, motivated by the
study of rank logic~\cite{DGHL09,GP15}, we consider gates that are
\emph{matrix-symmetric}. These are gates computing a Boolean function $f:
\{0,1\}^{A \times B} \rightarrow \{0,1\}$ which is invariant under arbitrary
permutations of $A$ and $B$. A typical example is the function computing the
rank of an $A \times B$ $0$-$1$ matrix. We show that the support theorem of
Anderson and Dawar can be shown to hold, even in the presence of
matrix-symmetric gates. This allows us to show that definability in the logic
fixed-point with rank ($\FPR$) is the same as decidability by
polynomially-uniform families of symmetric circuits with rank gates.
% Another extension of $\FPC$ that has received much attention in the
% descriptive complexity literature is the class $\CPTC$ of \emph{choiceless
% polynomial time with counting}. The work of~\cite{AD14} leaves open the
% question of obtaining a circuit characterization of this class and it was
% suggested there that this would require a relaxation of the notion of symmetry
% used in defining symmetric circuits. In particular, while $\CPTC$ is based on
% a machine model designed to preserve symmetry by forbidding choices, the
% polynomial time and space restrictions apply int he context of the symmetries
% fo a \emph{single input structure}. In this paper, we consider a relaxation of
% the notion of symmetric circuit intended to capture exactly this intuition.
% However, we are able to show that this does not yield a characterization of
% $\CPTC$. In particular, we show that no family in this class of relaxed
% symmetric circuits can distinguish odd from even Cai-F\"urer-Immerman graphs,
% something we know is do-able in $\CPTC$~\cite{DRR08}.

\end{document}

>>>>

\documentclass[../paper.tex]{subfiles}
\begin{document}
When we study algorithmic complexity we often study complexity classes defined
in terms of some restricted class of machines. Circuit complexity, where one
makes use of similarly restricted families of Boolean circuits rather than
machines, has long served as an alternative approach. These circuits models have
a natural attraction for researchers, as their simpler structure seems to make
them more amendable to analysis and their connections with machine models
provides an alternative avenue for better understanding fundamental problems in
complexity theory, including the $\PT$ vs $\NP$ problem.

Descriptive complexity instead concerns the characterisation of complexity
classes in terms of logics. The central open question in the field -- whether
there is a logical characterisation of the polynomial-time decidable properties
of finite relational structures -- essentially asks if there is a logic the
sentences of which describe all (and only) polynomial-time decidable
isomorphism-closed classes of graphs (i.e. every \emph{graph property} in
$\PT$). As in the algorithmic complexity case, the use of circuits in
descriptive complexity has proved useful, with Otto \cite{Otto1997} and
Denenberg et al.~\cite{DENENBERG1986216} developing circuit-based
characterisations for logics. These characterisation are given in terms of
\emph{symmetric circuits}, circuits that take in finite structures over some
universe and where permutations of the universe of the input structure are
reflected by automorphisms of the circuit.

% In this way symmetric circuits provide a natural class from which to develop
% circuit characterisations for logics, taking in abstract structures and
% ensuring that isomorphism (permutations) on the input structure are reflected
% in the circuit structure, which consequently implies that the function
% computed by the circuit is \emph{invariant} (i.e. computes a graph property)
% makes symmetric circuits a natural class from which to develop circuit
% characterisations of logics.

Anderson and Dawar \cite{AndersonD17} developed a similar
circuit-characterisation for $\FPC$, proving that families of P-uniform families
of symmetric circuits defined over a complete Boolean basis with majority gates
completely characterises \emph{fixed-point logic with counting} ($\FPC$).

While it is known that $\FPC$ does not capture $\PT$ \cite{Cau1992}, $\FPC$ does
capture $\PT$ over many classes of structures \cite{Grohe2010} and many natural
$\PT$ problems are expressible in $\FPC$ (e.g.\ Maximum Matching
\cite{Anderson2013}). While $\FPC$ is thus an important logic of interest for
its own sake, it also plays a role as a logic from which other important logics
of study are defined. The particular logic of interest in this paper is
\emph{fixed-point logic with rank} ($\FPR$), introduced by Dawar et
al.~\cite{Dawar09logicswith}, and is defined as an extension of $\FPC$ that
includes operators that can define the rank of a definable matrix over a fixed
finite field. This extension was prompted by the work of Atserias et
al.~\cite{Atserias20091666}, who had shown that that some of the most difficult
to express problems in $\PT$ (that are not expressible in $\FPC$) were reducible
to the general problem of solving systems of linear equations over prime fields.
Indeed, problems of this sort have remained the primary source of
difficult-to-express problems in descriptive complexity, motivating the modern
study of $\FPR$.

It is worth noting the logic defined by Dawar et al.\cite{} defines for each
prime a separate operator that computes the rank of the matrix over that prime.
It was shown by Gr\"{a}del and Pakusa \cite{GradelP15a} that this logic does not
capture $\PT$. Instead, they present a strictly more expressive alternative rank
logic which includes only a single rank operator that takes in both the matrix
and a specific prime over which to compute the rank as input. In this paper when
we refer to $\FPR$ we are referring to this more expressive version of
Gra\"{a}del and Pakusa.

In this paper we develop a circuit characterisation of $\FPR$ in terms of
symmetric circuits that have gates that compute rank. Concretely, a circuit
$C_n$ for structures of size $n$ is a directed acyclic graph (DAG) whose
internal gates are marked by elements of some basis. This basis consists of
operations or, more concretely, Boolean functions such that a gate $g$ in $C_n$
with $k$ input gates is labelled by a function $f: \{0,1\}^A \rightarrow
\{0,1\}$, where $\vert A \vert = k$. When evaluating the circuit, $g$ is
computed by evaluating $f$ with the input given by the binary string formed by
evaluating the gates input to $g$, implicitly mapping $A$ to the input gates of
$g$. However, the output of $f$ in general depends on the order (or structure)
of the binary string taken as input, and as such the evaluation of $g$ depends
on which gates are assigned to each element of $A$.

This point is often ignored in circuit complexity by ensuring that functions in
the Boolean basis are \emph{symmetric}, i.e. depend only on the number of 1's in
it's input, rendering the particular assignment from $A$ to the input gates
irrelevant. This is the approach taken by Anderson and Dawar \cite{AndersonD17}.
However, functions that compute an operation like rank, i.e. a function that
takes in a matrix with Boolean entries and outputs 1 iff that matrix, when
interpreted in $\mathbb{F}_p$, has rank at least $r$, certainly depend on how
the input is formed into a matrix, and as such is not a symmetric function.

We introduce the notion of a \emph{structured function}, which allows us to both
index inputs appropriately and introduce appropriate generalisations of symmetry
depending on the structure of the input. Each such function $F$ is uses a many-sorted
structure, with some signature $\tau_F$, in order to index the input. This
definition allows us to generalise the notion of symmetry to a particular
labelling, and so introduce the notion of a \emph{$\tau$-symmetric functions}.

We also introduce a generalised symmetric circuit allows, which allows for
$\tau$-symmetric functions in the Boolean basis, not just symmetric ones. The
circuit also provides a function $L$ that maps to each internal gate $g$ to a
labelling on its input corresponding the labelling on the input of the Boolean
function that labels this gate.

\begin{remark}
  I'm not sure how much detail to go into here, this could be a major
  digression.
\end{remark}

% That is, eacFor example, if we should like The circuit $C_n$ associates with
% each internal gate $g$ a structure $\dom (g)$, and each gate input a sequence of
% elements from each sort $\dom(g)$ in accordance with the sorts of the

% of definition includes an additional function which assosiates for each gate $g$
% a

% require that the Basis over which the circuit is defined are symemtric.. We
% begin by generalising the notion of a symmetric circuit. A symmetric circuit, as
% presented in \cite{AndersonD17}, and in many parts of complexity theory,
% supposes that the in Boolean function that labels in the gate is
% \emph{symmetric} that labels a particular gate We first generalise the notion of
% a symmetric circuit so as to allow for gate

% In this general notion of a symmetric circuit, a Boolean basis consists not just
% of a set of Boolean functions, but a set of \emph{structured functions} -- a
% function indexed A symmetric $\tau$-circuits is defined for some universe size
% $n$ and Boolean basis $\mathbb{B}$ of Boolean functions. Each Boolean function
% in $\mathbb{B}$ is a \emph{structured Boolean function} in the sense that

% In this abstract we review the work of Anderson and Dawar on symmetric
% circuits, and extend their analysis by broadening the basis of functions they
% consider to include so-called `matrix-symmetric' functions. This class of
% functions includes important non-symmetric functions, e.g.\ the rank of a
% matrix, and this allows us to study symmetric circuits with gates that compute
% rank. Our main result is a circuit characterisation for FPR. In particular, we
% show that P-uniform families of symmetric circuits with rank gates
% characterise FPR. This requires a significant advance on the methods
% of~\cite{AndersonD17}.
\end{document}
